---
title: "Maryland Lobbying Event Spending by Committee"
author: "Zaka Hossain"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

This analysis compares lobbying event spending by committee for two reporting periods:

- **2024-2025:** November 2024 – October 2025
- **2025-2026:** November 2025 – October 2026

Data source: Maryland State Ethics Commission

## Load Libraries

```{r libraries}
library(tidyverse)
library(janitor)
```

## Load Data

Update the file paths below to match your CSV locations.

```{r load-data}
events_2425 <- read_csv("Lobbying Event Reports - event_reports_nov2024-oct2025.csv") %>%
  clean_names()

events_2526 <- read_csv("event_reports_nov2025-oct2026 - event_reports_nov2025-oct2026.csv") %>%
  clean_names()
```

## Define Committee Patterns

```{r committee-patterns}
committees <- tribble(
  ~committee, ~pattern,
  "House ENT", "Environment and Transportation",
  "Senate EEE", "Education, Energy, and the Environment",
  "House Appropriations", "House Appropriations",
  "Senate Budget & Tax", "Budget and Taxation",
  "House Ways & Means", "Ways and Means",
  "Senate Finance", "Senate Finance",
  "House Economic Matters", "Economic Matters",
  "House Judiciary", "House Judiciary",
  "Senate Judicial Proceedings", "Judicial Proceedings",
  "General Assembly", "General Assembly"
)
```

## Function to Calculate Committee Spending

```{r spending-function}
get_committee_spending <- function(df, period_label) {
  
  results <- committees %>%
    mutate(
      total_spent = map_dbl(pattern, function(p) {
        df %>%
          filter(str_detect(groups_invited, regex(p, ignore_case = TRUE))) %>%
          summarize(total = sum(total_cost, na.rm = TRUE)) %>%
          pull(total)
      }),
      num_events = map_int(pattern, function(p) {
        df %>%
          filter(str_detect(groups_invited, regex(p, ignore_case = TRUE))) %>%
          nrow()
      }),
      period = period_label
    ) %>%
    select(committee, total_spent, num_events, period)
  
  return(results)
}
```

## Calculate Spending by Period

```{r calculate-spending}
spending_2425 <- get_committee_spending(events_2425, "2024-2025")
spending_2526 <- get_committee_spending(events_2526, "2025-2026")
```

## Year-over-Year Comparison

```{r comparison}
committee_comparison <- bind_rows(spending_2425, spending_2526) %>%
  pivot_wider(
    names_from = period,
    values_from = c(total_spent, num_events)
  ) %>%
  mutate(
    change_pct = round((`total_spent_2025-2026` - `total_spent_2024-2025`) / `total_spent_2024-2025` * 100, 1)
  ) %>%
  arrange(desc(`total_spent_2025-2026`))

knitr::kable(committee_comparison, 
             caption = "Committee Spending Comparison",
             format.args = list(big.mark = ","))
```

## Long Format for Visualization

```{r long-format}
committee_long <- bind_rows(spending_2425, spending_2526) %>%
  arrange(desc(total_spent))

knitr::kable(committee_long, 
             caption = "Committee Spending by Period",
             format.args = list(big.mark = ","))
```

## Export CSVs

```{r export}
write_csv(committee_comparison, "output_committee_comparison.csv")
write_csv(committee_long, "output_committee_spending_long.csv")
```

## Summary

```{r summary}
# Total spending by period
totals <- bind_rows(spending_2425, spending_2526) %>%
  group_by(period) %>%
  summarize(
    total_events = sum(num_events),
    total_spending = sum(total_spent)
  )

knitr::kable(totals, 
             caption = "Overall Totals by Period",
             format.args = list(big.mark = ","))
```
