---
title: "Maryland Lobbying Dinners: Data Analysis for CNS Story"
author: "Zaka Hossain, Capital News Service"
date: "2026-04-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
NOTE: This file has two jobs:
  1. Fact-check the dinner/reception figures in our CNS story on lobbying
  2. Generate CSV + chart outputs for data visualizations

SCOPE: Dinner and reception data ONLY. My colleague, Quinn Muller is handling the lobbying compensation/expenditure data (lobbyist salaries, client fees, etc.) separately.

DATA: Maryland State Ethics Commission lobbying event reports. These are disclosures lobbyists file when they host a dinner or reception for legislators — they report who attended, what committees were invited, and the total cost. Under Maryland law, regulated lobbyists who host a dinner or reception for a qualified legislative group — defined as all members of the General Assembly, either chamber, any standing committee or a recognized county or regional delegation — must file an expenditure report with the Commission within 14 days of the event. If final costs are not yet known, lobbyists may estimate and amend later. Capital News Service analyzed all event reports filed for the 2025 and 2026 regular sessions of the Maryland General Assembly. To account for events hosted in the days immediately surrounding each session, CNS included reports from the week of Jan. 6 through the week of Apr. 11, 2025, and from the week of Jan. 12 through the week of Apr. 17, 2026. Data for the 2025–2026 reporting year was retrieved April 27, 2026.
---

## Overview

This analysis uses Maryland State Ethics Commission lobbying event reports to:

- Check the dinner-related figures in our story against raw data
- Produce visualization-ready outputs for Datawrapper or Flourish

**Data sources:**

`Lobbying Event Reports - event_reports_nov2024-oct2025.csv` | Nov 2024 – Oct 2025 | Full year |
`event_reports_nov2025-oct2026 - event_reports_nov2025-oct2026_updated_04272026.csv` | Nov 2025 – Oct 2026 | Updated Apr 27, 2026 |

All output files are saved to `rstudio_output/`.

---

## Setup

```{r libraries}
# Core packages — install any missing ones with install.packages("packagename")
library(tidyverse)   # dplyr, ggplot2, readr, stringr, tidyr, purrr
library(janitor)     # clean_names() for standardizing column headers
library(lubridate)   # date parsing (mdy, ymd, etc.)
library(scales)      # dollar_format() for chart axes
library(knitr)       # kable() for summary tables

# Create output directory if it doesn't exist
if (!dir.exists("rstudio_output")) dir.create("rstudio_output")
```

---

## Load and Inspect Data

```{r load-data}
# 2024-2025 reporting year
events_2425 <- read_csv(
  "Lobbying Event Reports - event_reports_nov2024-oct2025.csv",
  show_col_types = FALSE
) %>%
  clean_names() %>%
  mutate(event_date_parsed = mdy(event_date))

# 2025-2026 reporting year (April 27 update)
# NOTE TO SELF: There are two 2025-26 files in this folder. The OLD one is
# "event_reports_nov2025-oct2026 - event_reports_nov2025-oct2026.csv"
# (used by the Python scripts). Always use the _updated_04272026 version.
events_2526 <- read_csv(
  "event_reports_nov2025-oct2026 - event_reports_nov2025-oct2026_updated_04272026.csv",
  show_col_types = FALSE
) %>%
  clean_names() %>%
  mutate(event_date_parsed = mdy(event_date))

# Quick sanity check
cat("2024-2025:", nrow(events_2425), "events\n")
cat("2025-2026:", nrow(events_2526), "events\n")
cat("\nColumns:", paste(names(events_2526), collapse = ", "), "\n")

# Flag any dates that failed to parse — these indicate data entry errors
cat("\nUnparsed dates in 2024-25:", sum(is.na(events_2425$event_date_parsed)), "\n")
cat("Unparsed dates in 2025-26:", sum(is.na(events_2526$event_date_parsed)), "\n")

# Show the bad rows so we can investigate
if (any(is.na(events_2526$event_date_parsed))) {
  cat("\nBad dates in 2025-26 data:\n")
  events_2526 %>%
    filter(is.na(event_date_parsed)) %>%
    select(employer, event_date, groups_invited) %>%
    print()
}

# Range check: mdy() parses two-digit years without error (e.g. "01/10/02" →
# 2002-01-10), so NA-only checks miss implausible dates. Flag anything outside
# each file's expected reporting window.
oor_2425 <- events_2425 %>%
  filter(event_date_parsed < ymd("2024-11-01") | event_date_parsed > ymd("2025-10-31"))
cat("\nOut-of-range dates in 2024-25 data (expect Nov 2024 – Oct 2025):", nrow(oor_2425), "record(s)\n")
print(oor_2425 %>% select(employer, event_date, event_date_parsed, total_cost))

oor_2526 <- events_2526 %>%
  filter(event_date_parsed < ymd("2025-11-01") | event_date_parsed > ymd("2026-10-31"))
cat("\nOut-of-range dates in 2025-26 data (expect Nov 2025 – Oct 2026):", nrow(oor_2526), "record(s)\n")
print(oor_2526 %>% select(employer, event_date, event_date_parsed, total_cost))
```

---

## Define Key Variables

```{r definitions}
# Maryland General Assembly session dates
# Using Monday of the week the session opens through Friday of the week it closes.
# This is more inclusive than exact dates — captures any events in the same week
# as the session opener/closer without requiring precise date matching.

# 2025: session opened Jan 8 (Wed) → window Jan 6 (Mon); closed Apr 7 (Mon) → window Apr 11 (Fri)
# 2026: session opened Jan 14 (Wed) → window Jan 12 (Mon); closed Apr 13 (Mon) → window Apr 17 (Fri)
session_2025_start <- ymd("2025-01-06")
session_2025_end   <- ymd("2025-04-11")

session_2026_start <- ymd("2026-01-12")
session_2026_end   <- ymd("2026-04-17")

# Committee name patterns for energy-related committees
# NOTE: Senate EEE was renamed from "Education, Health, and Environmental Affairs"
# in January 2023 (Senate President Ferguson). Energy/utilities portfolio moved
# from Senate Finance at the same time. Both datasets start after Jan 2023, so the
# old name will not appear in our data.
ent_pattern         <- "Environment and Transportation"
eee_pattern         <- "Education, Energy, and the Environment"
energy_cmte_pattern <- paste0(ent_pattern, "|", eee_pattern)

# Gas and electric utility companies — partial string match catches full names
# e.g., "BGE" catches "Baltimore Gas and Electric Company"
utility_pattern <- paste(
  c("BGE", "Baltimore Gas and Electric",
    "Constellation",
    "Pepco",
    "Washington Gas",
    "Choptank Electric",
    "Chesapeake Utilities",
    "Delmarva Power"),
  collapse = "|"
)

# Helper: standardize utility names to short versions for charts
standardize_utility <- function(name) {
  case_when(
    str_detect(name, regex("BGE|Baltimore Gas and Electric", ignore_case = TRUE)) ~ "BGE",
    str_detect(name, regex("Constellation",                  ignore_case = TRUE)) ~ "Constellation",
    str_detect(name, regex("Pepco",                          ignore_case = TRUE)) ~ "Pepco",
    str_detect(name, regex("Washington Gas",                 ignore_case = TRUE)) ~ "Washington Gas",
    str_detect(name, regex("Choptank",                       ignore_case = TRUE)) ~ "Choptank Electric",
    str_detect(name, regex("Chesapeake Util",                ignore_case = TRUE)) ~ "Chesapeake Utilities",
    str_detect(name, regex("Delmarva",                       ignore_case = TRUE)) ~ "Delmarva Power",
    TRUE ~ name
  )
}

# Pre-filter: all events during the 2026 session only
session_2026 <- events_2526 %>%
  filter(event_date_parsed >= session_2026_start,
         event_date_parsed <= session_2026_end)

# Pre-filter: ENT/EEE events in each year
ent_eee_2026 <- session_2026 %>%
  filter(str_detect(groups_invited, regex(energy_cmte_pattern, ignore_case = TRUE)))

ent_eee_2425 <- events_2425 %>%
  filter(str_detect(groups_invited, regex(energy_cmte_pattern, ignore_case = TRUE)))

cat("2026 session events:", nrow(session_2026), "\n")
cat("ENT/EEE events (2026 session):", nrow(ent_eee_2026), "\n")
cat("ENT/EEE events (2024-2025 full year):", nrow(ent_eee_2425), "\n")

# Confirm the pre-2023 committee name does not appear in either dataset
old_name_2425 <- events_2425 %>%
  filter(str_detect(groups_invited, regex("Education, Health, and Environmental Affairs", ignore_case = TRUE)))

old_name_2526 <- events_2526 %>%
  filter(str_detect(groups_invited, regex("Education, Health, and Environmental Affairs", ignore_case = TRUE)))

cat("\nOld EEE name in 2024-25 data:", nrow(old_name_2425), "rows\n")
cat("Old EEE name in 2025-26 data:", nrow(old_name_2526), "rows\n")
if (nrow(old_name_2425) > 0 | nrow(old_name_2526) > 0) {
  cat("NOTE: Old name found — review these rows and update energy_cmte_pattern.\n")
}
```

---

# PART I — FACT-CHECKS

*Ordered by appearance in the story. Each section quotes the story sentence being verified, then shows what the data says.*

---

## FC-1: Cornerstone Government Affairs, April 8 Dinner at Lewnes'

> *"The firm, Cornerstone Government Affairs, picked up the tab for this April 8 dinner for members of the House Appropriations Committee and Senate Budget and Taxation Committee. The bill closed out at $9,275.11, according to the Maryland State Ethics Commission."*

```{r fc1-cornerstone}
# Search on both the employer field AND the organization_firm field (the lobbying firm)
# because Cornerstone may file under a client's name as employer
cornerstone_events <- events_2526 %>%
  filter(
    str_detect(employer,          regex("Cornerstone", ignore_case = TRUE)) |
    str_detect(organization_firm, regex("Cornerstone", ignore_case = TRUE))
  ) %>%
  select(event_date, event_date_parsed, employer, organization_firm,
         total_cost, groups_invited, sponsors)

cat("All Cornerstone-related events in 2025-26 data:\n")
print(cornerstone_events)

# Specifically the April 8 event
cornerstone_apr8 <- cornerstone_events %>%
  filter(event_date_parsed == ymd("2026-04-08"))

if (nrow(cornerstone_apr8) > 0) {
  cat("\nApril 8, 2026 event:\n")
  print(cornerstone_apr8)
} else {
  cat("\nNOTE FOR STORY: The Cornerstone April 8 dinner is NOT yet in the dataset.\n")
  cat("The $9,275.11 figure came from a physical receipt — verify it matches\n")
  cat("what Cornerstone reports when filings are complete.\n")
}
```

---

## FC-2: Event Frequency During the 2026 Session

> *"Closed-door dinners and private receptions for committees and the General Assembly happened nearly every two days during the 2026 legislative session that ended in mid-April, according to calendars sent to all legislators, obtained and analyzed by CNS."*

```{r fc2-frequency}
session_days       <- as.numeric(session_2026_end - session_2026_start) + 1
unique_event_dates <- n_distinct(session_2026$event_date_parsed)
total_events_2026  <- nrow(session_2026)

cat("2026 session: January 12 – April 17 (week-of-session window),", session_days, "days\n")
cat("Days with at least one dinner/reception:", unique_event_dates, "\n")
cat("Total events filed:", total_events_2026, "\n")
cat("Average: one event day every", round(session_days / unique_event_dates, 1), "days\n")
```

---

## FC-3: Total Spending, Year Ending October 2025

> *"The most recent data available shows spending on those gatherings rose to $2.3 million for the year ending October 2025, according to a CNS analysis of the commission's lobbying records."*

```{r fc3-annual-total}
# Note: Using 2024-2025 data because it is a complete reporting year.
# The 2025-2026 data is only through April 27, 2026.
total_2425 <- events_2425 %>%
  summarize(
    total_events   = n(),
    total_spending = sum(total_cost, na.rm = TRUE)
  )

cat("Year ending October 2025\n")
cat("Total events:  ", total_2425$total_events, "\n")
cat("Total spending: $", format(total_2425$total_spending, big.mark = ",", nsmall = 2), "\n")
```

---

## FC-3b: Total Spending, Year Ending October 2026 (Partial)

> *"The most recent data available shows spending on those gatherings rose to $2.3 million for the year ending October 2025, according to a CNS analysis of the commission's lobbying records."*

*Note: The 2025-2026 reporting year is not yet complete — data retrieved April 27, 2026. This is for comparison only.*

```{r fc3b-annual-total-2526}
total_2526 <- events_2526 %>%
  summarize(
    total_events   = n(),
    total_spending = sum(total_cost, na.rm = TRUE)
  )

cat("Year ending October 2026 (partial — retrieved April 27, 2026)\n")
cat("Total events:  ", total_2526$total_events, "\n")
cat("Total spending: $", format(total_2526$total_spending, big.mark = ",", nsmall = 2), "\n")

cat("\nYear-over-year comparison:\n")
cat("2024-2025: $", format(total_2425$total_spending, big.mark = ",", nsmall = 2), "\n")
cat("2025-2026: $", format(total_2526$total_spending, big.mark = ",", nsmall = 2), "(partial)\n")
cat("Change so far: $", format(total_2526$total_spending - total_2425$total_spending, big.mark = ",", nsmall = 2), "\n")
```

---

## FC-4: Utility Spending — All Events During 2026 Session

> *"Utility companies that do business in Maryland have reported spending $60,638 in total, or about $6,064 per event, during the 2026 legislative session, on committee and General Assembly dinners and receptions while these same bodies were deliberating over policies that affect the companies' profits and consumers' bills, according to the CNS analysis."*

```{r fc4-utility-all-events-2026}
utility_all_2026 <- session_2026 %>%
  filter(str_detect(employer, regex(utility_pattern, ignore_case = TRUE)))

utility_all_total_2026 <- utility_all_2026 %>%
  summarize(
    total_events   = n(),
    total_spending = sum(total_cost, na.rm = TRUE),
    avg_per_event  = mean(total_cost, na.rm = TRUE)
  )

cat("Utility spending on ALL committee events, 2026 session:\n")
cat("Events:        ", utility_all_total_2026$total_events, "\n")
cat("Total:          $", format(utility_all_total_2026$total_spending, big.mark = ",", nsmall = 2), "\n")
cat("Avg per event:  $", format(utility_all_total_2026$avg_per_event, big.mark = ",", nsmall = 2), "\n")

cat("\nBreakdown by utility:\n")
utility_all_2026 %>%
  group_by(employer) %>%
  summarize(
    events = n(),
    total  = sum(total_cost, na.rm = TRUE)
  ) %>%
  arrange(desc(total)) %>%
  print()
```

---

## FC-5: Utility ENT/EEE Spending and Total ENT/EEE Spending, 2026 Session

> *"During the 2026 legislative session, the utility companies BGE, Pepco, Washington Gas, Choptank Electric and Constellation reported spending $55,142 treating the House ENT and the Senate EEE committees — the two committees with jurisdiction over their businesses — to dinners and receptions, according to the CNS analysis. That figure represents nearly a third of the $172,878 lobbyists spent on meals and receptions for those two committees during the session."*

```{r fc5-utility-ent-eee-and-total-2026}
# Utility spending on ENT/EEE only
utility_ent_eee_2026 <- ent_eee_2026 %>%
  filter(str_detect(employer, regex(utility_pattern, ignore_case = TRUE)))

utility_ent_eee_total_2026 <- utility_ent_eee_2026 %>%
  summarize(
    total_events   = n(),
    total_spending = sum(total_cost, na.rm = TRUE),
    avg_per_event  = mean(total_cost, na.rm = TRUE)
  )

# Total ENT/EEE spending from ALL employers (the $172,878 figure)
ent_eee_all_total_2026 <- ent_eee_2026 %>%
  summarize(
    total_events   = n(),
    total_spending = sum(total_cost, na.rm = TRUE)
  )

# 2026 session total for context
session_2026_total <- session_2026 %>%
  summarize(
    total_events   = n(),
    total_spending = sum(total_cost, na.rm = TRUE)
  )

cat("=== UTILITY spending on ENT/EEE, 2026 session ===\n")
cat("Events:        ", utility_ent_eee_total_2026$total_events, "\n")
cat("Total:          $", format(utility_ent_eee_total_2026$total_spending, big.mark = ",", nsmall = 2), "\n")
cat("Avg per event:  $", format(utility_ent_eee_total_2026$avg_per_event, big.mark = ",", nsmall = 2), "\n")

cat("\n=== ALL employers on ENT/EEE, 2026 session ===\n")
cat("Events:        ", ent_eee_all_total_2026$total_events, "\n")
cat("Total:          $", format(ent_eee_all_total_2026$total_spending, big.mark = ",", nsmall = 2), "\n")

cat("\nUtility share of ENT/EEE spending:",
    round(utility_ent_eee_total_2026$total_spending / ent_eee_all_total_2026$total_spending * 100, 1), "%\n")

cat("\n=== 2026 session total (all employers, all committees) ===\n")
cat("Events:  ", session_2026_total$total_events, "\n")
cat("Total:   $", format(session_2026_total$total_spending, big.mark = ",", nsmall = 2), "\n")

cat("\nENT/EEE share of total session spending:",
    round(ent_eee_all_total_2026$total_spending / session_2026_total$total_spending * 100, 1), "%\n")

cat("\nEvent-by-event utility breakdown:\n")
utility_ent_eee_2026 %>%
  select(event_date, employer, total_cost, groups_invited) %>%
  arrange(desc(total_cost)) %>%
  print()
```

---

# PART II — ADDITIONAL ANALYSIS

*Supporting calculations not directly cited in the story but used to verify numbers, check sourcing, or provide background context.*

---

## Extra-1: Total Spending, Year Ending October 2026 (Partial)

```{r extra1-annual-total-2526}
total_2526 <- events_2526 %>%
  summarize(
    total_events   = n(),
    total_spending = sum(total_cost, na.rm = TRUE)
  )

cat("Year ending October 2026 (partial — retrieved April 27, 2026)\n")
cat("Total events:  ", total_2526$total_events, "\n")
cat("Total spending: $", format(total_2526$total_spending, big.mark = ",", nsmall = 2), "\n")
```

---

## Extra-2: 2025 Session Total and ENT/EEE Share

```{r extra2-session-2025-total}
session_2025 <- events_2425 %>%
  filter(event_date_parsed >= session_2025_start,
         event_date_parsed <= session_2025_end)

session_2025_total <- session_2025 %>%
  summarize(
    total_events   = n(),
    total_spending = sum(total_cost, na.rm = TRUE)
  )

ent_eee_2025_session <- session_2025 %>%
  filter(str_detect(groups_invited, regex(energy_cmte_pattern, ignore_case = TRUE)))

ent_eee_2025_total <- ent_eee_2025_session %>%
  summarize(
    total_events   = n(),
    total_spending = sum(total_cost, na.rm = TRUE)
  )

cat("2025 session total\n")
cat("Events:  ", session_2025_total$total_events, "\n")
cat("Total:   $", format(session_2025_total$total_spending, big.mark = ",", nsmall = 2), "\n")

cat("\nENT/EEE share of 2025 session\n")
cat("Events:  ", ent_eee_2025_total$total_events, "\n")
cat("Total:   $", format(ent_eee_2025_total$total_spending, big.mark = ",", nsmall = 2), "\n")
cat("Share:   ", round(ent_eee_2025_total$total_spending / session_2025_total$total_spending * 100, 1), "%\n")
```

---

## Extra-3: Utility Spending — All Events During 2025 Session

```{r extra3-utility-all-events-2025}
utility_all_2025 <- session_2025 %>%
  filter(str_detect(employer, regex(utility_pattern, ignore_case = TRUE)))

utility_all_total_2025 <- utility_all_2025 %>%
  summarize(
    total_events   = n(),
    total_spending = sum(total_cost, na.rm = TRUE),
    avg_per_event  = mean(total_cost, na.rm = TRUE)
  )

cat("Utility spending on ALL events, 2025 session:\n")
cat("Events:        ", utility_all_total_2025$total_events, "\n")
cat("Total:          $", format(utility_all_total_2025$total_spending, big.mark = ",", nsmall = 2), "\n")
cat("Avg per event:  $", format(utility_all_total_2025$avg_per_event, big.mark = ",", nsmall = 2), "\n")

cat("\nBreakdown by utility:\n")
utility_all_2025 %>%
  group_by(employer) %>%
  summarize(
    events = n(),
    total  = sum(total_cost, na.rm = TRUE)
  ) %>%
  arrange(desc(total)) %>%
  print()
```

---

## Extra-4: BGE, Pepco, Washington Gas & Constellation on ENT/EEE — 2024-2025

*Used to investigate the original $70,000 claim, which turned out to be total utility spending on all events (not ENT/EEE specific). See data notes.*

```{r extra4-four-utilities-ent-eee-2425}
four_utilities_pattern <- "BGE|Baltimore Gas and Electric|Pepco|Washington Gas|Constellation"

# Full year
four_utilities_ent_eee_2425 <- ent_eee_2425 %>%
  filter(str_detect(employer, regex(four_utilities_pattern, ignore_case = TRUE)))

four_utilities_total <- four_utilities_ent_eee_2425 %>%
  summarize(
    total_events   = n(),
    total_spending = sum(total_cost, na.rm = TRUE)
  )

cat("BGE, Pepco, Washington Gas & Constellation — ENT/EEE, 2024-2025 full year:\n")
cat("Events:  ", four_utilities_total$total_events, "\n")
cat("Total:    $", format(four_utilities_total$total_spending, big.mark = ",", nsmall = 2), "\n")

# Session window only
four_utilities_ent_eee_2425_session <- four_utilities_ent_eee_2425 %>%
  filter(event_date_parsed >= session_2025_start,
         event_date_parsed <= session_2025_end)

four_utilities_session_total <- four_utilities_ent_eee_2425_session %>%
  summarize(
    total_events   = n(),
    total_spending = sum(total_cost, na.rm = TRUE)
  )

cat("\nSession window only (Jan 6 – Apr 11, 2025):\n")
cat("Events:  ", four_utilities_session_total$total_events, "\n")
cat("Total:    $", format(four_utilities_session_total$total_spending, big.mark = ",", nsmall = 2), "\n")
```

---

## Extra-5: BGE, Pepco, Washington Gas & Constellation on ENT/EEE — 2025-2026

```{r extra5-four-utilities-ent-eee-2526}
ent_eee_2526 <- events_2526 %>%
  filter(str_detect(groups_invited, regex(energy_cmte_pattern, ignore_case = TRUE)))

four_utilities_ent_eee_2526 <- ent_eee_2526 %>%
  filter(str_detect(employer, regex(four_utilities_pattern, ignore_case = TRUE)))

four_utilities_total_2526 <- four_utilities_ent_eee_2526 %>%
  summarize(
    total_events   = n(),
    total_spending = sum(total_cost, na.rm = TRUE)
  )

cat("BGE, Pepco, Washington Gas & Constellation\n")
cat("ENT/EEE events, full year Nov 2025 – Oct 2026 (partial, retrieved Apr 27):\n")
cat("Events:  ", four_utilities_total_2526$total_events, "\n")
cat("Total:   $", format(four_utilities_total_2526$total_spending, big.mark = ",", nsmall = 2), "\n")

cat("\nBreakdown by utility:\n")
four_utilities_ent_eee_2526 %>%
  group_by(employer) %>%
  summarize(
    events = n(),
    total  = sum(total_cost, na.rm = TRUE)
  ) %>%
  arrange(desc(total)) %>%
  print()
```

---

## Extra-6: All Utilities on Any Event — 2024-2025 Full Year

*This is what the original Python script calculated and mislabeled as "utility spending on ENT/EEE." The ~$70K figure came from this, not from ENT/EEE-specific filtering.*

```{r extra6-all-utilities-any-event-2425}
all_utilities_any_event_2425 <- events_2425 %>%
  filter(str_detect(employer, regex(utility_pattern, ignore_case = TRUE)))

all_utilities_any_event_2425 %>%
  group_by(employer) %>%
  summarize(
    events = n(),
    total  = sum(total_cost, na.rm = TRUE)
  ) %>%
  arrange(desc(total)) %>%
  print()

cat("\nTotal: $", format(sum(all_utilities_any_event_2425$total_cost, na.rm = TRUE), big.mark = ",", nsmall = 2), "\n")
```

---

## Extra-7: Every Utility Event on ENT/EEE — Both Years, Every Row

```{r extra7-utility-ent-eee-all-rows}
util_ent_eee_2425_all <- events_2425 %>%
  filter(str_detect(groups_invited, regex(energy_cmte_pattern, ignore_case = TRUE))) %>%
  filter(str_detect(employer, regex(utility_pattern, ignore_case = TRUE))) %>%
  mutate(year = "2024-2025") %>%
  select(year, event_date, employer, organization_firm, total_cost, groups_invited)

util_ent_eee_2526_all <- events_2526 %>%
  filter(str_detect(groups_invited, regex(energy_cmte_pattern, ignore_case = TRUE))) %>%
  filter(str_detect(employer, regex(utility_pattern, ignore_case = TRUE))) %>%
  mutate(year = "2025-2026") %>%
  select(year, event_date, employer, organization_firm, total_cost, groups_invited)

util_ent_eee_both <- bind_rows(util_ent_eee_2425_all, util_ent_eee_2526_all) %>%
  arrange(year, desc(total_cost))

cat("=== EVERY UTILITY EVENT ON ENT/EEE ===\n")
cat("2024-2025 events:", nrow(util_ent_eee_2425_all), "\n")
cat("2025-2026 events:", nrow(util_ent_eee_2526_all), "\n")
cat("Total events:", nrow(util_ent_eee_both), "\n\n")

print(util_ent_eee_both, n = Inf)

cat("\n2024-2025 total: $", format(sum(util_ent_eee_2425_all$total_cost, na.rm = TRUE), big.mark = ",", nsmall = 2), "\n")
cat("2025-2026 total: $", format(sum(util_ent_eee_2526_all$total_cost, na.rm = TRUE), big.mark = ",", nsmall = 2), "\n")
```

---

## Extra-8: Date Parsing Check — Utility Events

```{r extra8-date-parse-check}
# Confirm no utility events have unparsed dates that would be silently dropped
events_2526 %>%
  filter(str_detect(employer, regex(utility_pattern, ignore_case = TRUE))) %>%
  select(event_date, event_date_parsed, employer, total_cost) %>%
  arrange(event_date_parsed) %>%
  print(n = Inf)
```

---

# PART III — VISUALIZATIONS

*All charts are built in Datawrapper or Flourish. This section generates the CSVs and prints row previews for verification. All files saved to `rstudio_output/`.*

---

## VIZ-1: Who's Buying Dinner for Energy Lawmakers (Horizontal Bar)

**Tool:** Datawrapper
**Headline:** Who's buying dinner for Maryland's energy lawmakers?
**Subhead:** Engineering firms, farm groups and utilities led spending on House ENT and Senate EEE committee events in 2026
**Source:** Maryland State Ethics Commission
**Byline:** Zaka Hossain, Capital News Service
**Footnote:** Note: Rifkin Weiner Livingston and G.S. Proctor & Associates, Inc. are lobbying firms that do not identify the organizations they represent in their event reports filed with the Maryland State Ethics Commission. A portion of G.S. Proctor & Associates, Inc.'s spending could be on behalf of any of their clients, which include Pepco and Washington Gas.
**Colors:** Gas/Electric Utility = #d85a30 | Renewable Energy = #1d9e75 | Other Interest = #888780

```{r viz1-top-spenders}
renewable_pattern <- "solar|wind|renewable|MAREC|CHESSA|Chaberton|Advanced Energy"

# Check that Rifkin Weiner Livingston and G.S. Proctor appear in the data
# before using them in the footnote
cat("Footnote firm check:\n")
ent_eee_2026 %>%
  filter(str_detect(organization_firm,
                    regex("Rifkin|Proctor", ignore_case = TRUE))) %>%
  select(event_date, employer, organization_firm, total_cost) %>%
  print()

top_spenders_ent_eee <- ent_eee_2026 %>%
  mutate(
    category = case_when(
      str_detect(employer, regex(utility_pattern,   ignore_case = TRUE)) ~ "Gas/Electric Utility",
      str_detect(employer, regex(renewable_pattern, ignore_case = TRUE)) ~ "Renewable Energy",
      TRUE ~ "Other Interest"
    )
  ) %>%
  group_by(employer, category) %>%
  summarize(
    total_spent = sum(total_cost, na.rm = TRUE),
    num_events  = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(total_spent)) %>%
  slice_head(n = 15)

write_csv(top_spenders_ent_eee, "rstudio_output/output_top_spenders_ent_eee.csv")

cat("\nPreview (output_top_spenders_ent_eee.csv):\n")
print(top_spenders_ent_eee)
cat("Saved: rstudio_output/output_top_spenders_ent_eee.csv\n")
```

---

## VIZ-2: Utilities Ramp Up Spending on Energy Committees, 2025 vs. 2026 Session (Grouped Bar)

**Tool:** Datawrapper
**Headline:** Utility companies ramp up wining and dining in 2026
**Subhead:** Major utilities increased lobbying event spending on energy committees during the 2026 session compared to the 2025 session
**Source:** Maryland State Ethics Commission
**Byline:** Zaka Hossain, Capital News Service
**Footer:** 2025 session: week of Jan. 6 – week of Apr. 11. 2026 session: week of Jan. 12 – week of Apr. 17. Data as of April 27, 2026.
**Colors:** 2025 session = #888780 (gray) | 2026 session = #d85a30 (orange)
**Format for Datawrapper:** Wide — one row per utility, columns for each session

```{r viz2-utility-session-comparison}
# 2025 SESSION only — filter 2024-2025 data to session window, then ENT/EEE, then utilities
ent_eee_2025_session_viz <- events_2425 %>%
  filter(event_date_parsed >= session_2025_start,
         event_date_parsed <= session_2025_end) %>%
  filter(str_detect(groups_invited, regex(energy_cmte_pattern, ignore_case = TRUE))) %>%
  filter(str_detect(employer, regex(utility_pattern, ignore_case = TRUE)))

utility_session_2025 <- ent_eee_2025_session_viz %>%
  mutate(utility = standardize_utility(employer)) %>%
  group_by(utility) %>%
  summarize(`2025 session` = sum(total_cost, na.rm = TRUE), .groups = "drop")

# 2026 SESSION — already filtered in definitions chunk
utility_session_2026 <- utility_ent_eee_2026 %>%
  mutate(utility = standardize_utility(employer)) %>%
  group_by(utility) %>%
  summarize(`2026 session` = sum(total_cost, na.rm = TRUE), .groups = "drop")

# Wide format for Datawrapper
utility_comparison_wide <- full_join(utility_session_2025, utility_session_2026,
                                     by = "utility") %>%
  mutate(across(where(is.numeric), ~replace_na(., 0))) %>%
  arrange(desc(`2026 session`))

write_csv(utility_comparison_wide, "rstudio_output/output_utility_comparison.csv")

cat("Preview (output_utility_comparison.csv):\n")
print(utility_comparison_wide)
cat("Saved: rstudio_output/output_utility_comparison.csv\n")
```

---

## VIZ-3: Who Paid for Energy Committee Dinners (Sector Breakdown Donut)

**Tool:** Datawrapper
**Status:** May keep in story
**Headline:** Utilities accounted for nearly a third of energy committee dinner spending
**Subhead:** Gas and electric companies spent $55,142 of the $172,878 lobbyists reported on House ENT and Senate EEE events during the 2026 session
**Source:** Maryland State Ethics Commission
**Byline:** Zaka Hossain, Capital News Service
**Colors:** Gas/Electric Utility = #d85a30 | Renewable Energy = #1d9e75 | Other Interest = #888780

```{r viz3-sector-donut}
ent_eee_sector_breakdown <- ent_eee_2026 %>%
  mutate(
    sector = case_when(
      str_detect(employer, regex(utility_pattern,   ignore_case = TRUE)) ~ "Gas/Electric Utility",
      str_detect(employer, regex(renewable_pattern, ignore_case = TRUE)) ~ "Renewable Energy",
      TRUE ~ "Other Interest"
    )
  ) %>%
  group_by(sector) %>%
  summarize(total_spent = sum(total_cost, na.rm = TRUE), .groups = "drop") %>%
  mutate(percentage = round(total_spent / sum(total_spent) * 100, 1)) %>%
  arrange(desc(total_spent))

write_csv(ent_eee_sector_breakdown, "rstudio_output/output_ent_eee_sector_breakdown.csv")

cat("Preview (output_ent_eee_sector_breakdown.csv):\n")
print(ent_eee_sector_breakdown)
cat("Saved: rstudio_output/output_ent_eee_sector_breakdown.csv\n")
```

---

## VIZ-4: The Session Dinner Calendar (Scatter Timeline)

**Tool:** Flourish
**Headline:** The session dinner calendar
**Subhead:** Energy lobbyists hosted `r nrow(ent_eee_2026)` events for ENT and EEE members during the 2026 session as the Utility RELIEF Act advanced
**Source:** Maryland State Ethics Commission • Data as of April 27, 2026
**Byline:** Zaka Hossain, Capital News Service
**X-axis:** event_date | **Y-axis:** category (three horizontal bands) | **Size:** total_cost | **Color:** category
**Colors:** Gas/Electric Utility = #d85a30 | Renewable Energy = #1d9e75 | Other Interest = #888780
**Annotations (vertical lines):** ENT hearing :: 03/03/26 | House passes :: 03/17/26 | EEE hearing :: 03/25/26

```{r viz4-timeline}
timeline_ent_eee <- ent_eee_2026 %>%
  mutate(
    category = case_when(
      str_detect(employer, regex(utility_pattern,   ignore_case = TRUE)) ~ "Gas/Electric Utility",
      str_detect(employer, regex(renewable_pattern, ignore_case = TRUE)) ~ "Renewable Energy",
      TRUE ~ "Other Interest"
    ),
    target = case_when(
      str_detect(groups_invited, regex(ent_pattern, ignore_case = TRUE)) &
        str_detect(groups_invited, regex(eee_pattern, ignore_case = TRUE)) ~ "Both ENT & EEE",
      str_detect(groups_invited, regex(ent_pattern, ignore_case = TRUE))   ~ "House ENT only",
      TRUE                                                                  ~ "Senate EEE only"
    )
  ) %>%
  select(event_date, employer, organization_firm,
         total_cost, category, target, groups_invited) %>%
  arrange(event_date)

write_csv(timeline_ent_eee, "rstudio_output/output_ent_eee_timeline_2026.csv")

cat("Preview (output_ent_eee_timeline_2026.csv):\n")
print(timeline_ent_eee, n = Inf)
cat("Saved: rstudio_output/output_ent_eee_timeline_2026.csv\n")
```

---

## VIZ-5: Event Frequency During the 2026 Session (Bar)

**Tool:** Datawrapper
**Headline:** Lobbyists hosted a dinner or reception for legislators nearly every two days during the 2026 session
**Subhead:** Spending on lobbyist-funded meals and receptions for Maryland legislators fluctuated throughout the 90-day legislative session
**Source:** Maryland State Ethics Commission
**Byline:** Zaka Hossain, Capital News Service
**Footer:** Week of Jan. 12 – week of Apr. 17, 2026. Each bar represents one event day; height = total spending that day.

```{r viz5-frequency}
session_cal <- session_2026 %>%
  group_by(event_date_parsed) %>%
  summarize(
    num_events     = n(),
    total_spending = sum(total_cost, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  complete(
    event_date_parsed = seq.Date(session_2026_start, session_2026_end, by = "day"),
    fill = list(num_events = 0, total_spending = 0)
  ) %>%
  mutate(
    has_event       = num_events > 0,
    # Format date as string for Datawrapper
    date_label      = format(event_date_parsed, "%m/%d/%Y")
  ) %>%
  select(date_label, event_date_parsed, num_events, total_spending, has_event)

write_csv(session_cal, "rstudio_output/output_session_events_by_date.csv")

cat("Preview — event days only:\n")
session_cal %>%
  filter(has_event) %>%
  print(n = Inf)

cat("\nTotal event days:", sum(session_cal$has_event), "out of", nrow(session_cal), "session days\n")
cat("Saved: rstudio_output/output_session_events_by_date.csv\n")
```

---

# PART IV — SUMMARY

## Key Numbers: Story vs. Data

```{r summary-table}
# NOTE TO PROF: This table shows side-by-side what the story says versus
# what the data shows. Figures not in our dataset (e.g., the 1999 historical
# figure) come from published Ethics Commission annual reports.

cornerstone_note <- if (nrow(cornerstone_apr8) > 0) {
  paste0("$", format(cornerstone_apr8$total_cost[1], big.mark = ",", nsmall = 2), " — confirmed in data")
} else {
  "Not yet filed — verify when filings are complete"
}

summary_df <- tibble(
  `Story figure` = c(
    "Cornerstone dinner Apr 8: $9,275.11",
    "2026 session frequency: nearly every two days",
    "Year ending Oct 2025 total: $2.3 million",
    "Utility spending, 2026 session (all events): $60,638",
    "Avg cost per utility event: ~$6,064",
    "Utility spending on ENT/EEE, 2026 session: $55,142",
    "Total ENT/EEE spending, 2026 session: $172,878",
    "1999 receptions total: $684,958"
  ),
  `What the data shows` = c(
    cornerstone_note,
    paste0("Every ", round(session_days / unique_event_dates, 1), " days"),
    paste0("$", format(total_2425$total_spending, big.mark = ",", nsmall = 2)),
    paste0("$", format(utility_all_total_2026$total_spending, big.mark = ",", nsmall = 2),
           " (", utility_all_total_2026$total_events, " events)"),
    paste0("$", format(utility_all_total_2026$avg_per_event, big.mark = ",", nsmall = 2)),
    paste0("$", format(utility_ent_eee_total_2026$total_spending, big.mark = ",", nsmall = 2),
           " (", utility_ent_eee_total_2026$total_events, " events)"),
    paste0("$", format(ent_eee_all_total_2026$total_spending, big.mark = ",", nsmall = 2),
           " (", ent_eee_all_total_2026$total_events, " events)"),
    "From 1999 Ethics Commission annual report — not in our dataset"
  )
)

kable(summary_df,
      col.names = c("Story figure", "What the data shows"),
      caption = "Fact-check summary — confirm each figure before publication")
```

---

## Output Files Generated

```{r file-list}
cat("Files in rstudio_output/:\n")
list.files("rstudio_output/") %>%
  sort() %>%
  cat(sep = "\n")
```

---

## Data Notes

```{r data-notes}
cat("
DATA CAVEATS:

1. DATA CURRENCY: The 2025-2026 file was retrieved April 27, 2026 and
   captures events through Sine Die (April 13, 2026). 

2. MULTI-COMMITTEE EVENTS: When one event invites multiple committees
   (e.g., 'House Appropriations; Senate Budget and Taxation'), the full
   cost is attributed to each committee in our count. Costs are NOT split
   proportionally. Committee totals can therefore overlap.

3. UTILITY PATTERN MATCHING: We match on partial strings (e.g., 'BGE'
   catches 'Baltimore Gas and Electric Company'). Spot-check the employer
   column in output CSVs against actual filings if any numbers seem off.

5. CORNERSTONE DINNER: The $9,275.11 came from a physical receipt.
   It matches the filing in our dataset -- this isnt a caveat anymore.

6. BAD DATE RECORD: League of Women Voters of Howard County has event_date
   '01/10/02', which parses as 2002-01-10 — a likely two-digit year typo.
   The event costs $524. Because the parsed date falls outside all session
   and reporting-year windows, this record is excluded from every filter
   in this analysis and does not materially affect any totals. The
   out-of-range check in the load-data chunk flags it on every knit.
")
```

---

*Analysis by Zaka Hossain, Capital News Service. Generated `r format(Sys.Date(), "%B %d, %Y")`.*
